---
title: "Statistical Inference Course Project: Exploring the Central Limit Theorem"
author: "By Matthew Sedlar"
output: 
  html_document:
    keep_md: true
---


## Overview: In a few (2-3) sentences explain what is going to be reported on.

## Simulations

Let's start by setting the number of simulations we want to run as well as the number of exponentials and rate for each simulation.

```{r}

nosims <- 1000
nexps <- 40
rate <- 0.2

```

Let's also not forget to set.seed, so that our outcome is the same every time we run the script.

```{r}

# 13 is my lucky number
set.seed(13)

```

I'm going to start off by running 1,000 simulations on the exponentials using R's replicate function. For each simulation, I'm finding the mean and storing it in a numeric vector called sims.

```{r}

sims <- replicate(nosims, mean(rexp(nexps,rate)))

```

Let's plot the sims object using ggplot's histogram and density functions to get a sense of our distribution.

```{r}

library(ggplot2)

ggplot(data.frame(sims), aes(sims)) + 
  geom_histogram(aes(y = ..density..)) +
  geom_density()


```

## Sample Mean versus Theoretical Mean: Include figures with titles. In the figures, highlight the means you are comparing. Include text that explains the figures and what is shown on them, and provides appropriate numbers.

## Sample Variance versus Theoretical Variance: Include figures (output from R) with titles. Highlight the variances you are comparing. Include text that explains your understanding of the differences of the variances.

## Distribution: Via figures and text, explain how one can tell the distribution is approximately normal.
