---
title: 'Statistical Inference Course Project: Exploring the Central Limit Theorem'
author: "By Matthew Sedlar"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---


## Overview

This report explores the Central Limit Theorem using a set of randomly generated exponentials. It will look at the sample mean and how it relates to the theoretical mean, the sample variance and how it relates to the theoretical variance, and whether the resulting distribution approximates normal.

## Simulations

I'm going to run 1,000 simulations on 40 randomly generated exponentials with a rate of 0.2 using R's replicate function. Once I have the simulation results in the form of a matrix, I will run the mean function on the matrix using apply, then store the results in a data frame along with the number of simulations. (See Appendix A for the full code). 

```{r echo=FALSE}

nosims <- 1000
nexps <- 40
rate <- 0.2

# using set.seed so the results can be replicated
set.seed(13)

# 1000 simulations on 40 random exponentials
sims <- replicate(nosims,rexp(nexps,rate))

# storing the clt of those simulations in a data frame
# with the size of the simulations
simsdf <- data.frame(averages = apply(sims,2, mean),
                     size = factor("1000"))

```

## Sample Mean versus Theoretical Mean

Let's examine the distribution for our simulation (See Appendix B for a plot of the distribution). Our sample mean is: 

```{r echo=FALSE}

samplemean <- mean(simsdf$averages)
samplemean

```

We know given our rate (0.2) that the theoretical mean is 5 based on the following formula:

$$E[X]=\frac{1}{\lambda}={\beta}$$

We know from the Central Limit Theorem that the sample mean from unbiased data approximates the population mean. To further illustrate this, I'm going to run another 5,000 simulations and then compare the two distributions (see Appendix C for code). 


```{r echo=FALSE, message=FALSE, fig.width=4.5, fig.height=3.5, fig.align='center'}

# simulate 5000 random exponentials
set.seed(13)
sims2 <- replicate(nosims*5,rexp(nexps,rate))

# creating a new data frame and then binding those rows to the original
sampledf <- data.frame(averages=c(
  apply(sims2,2, mean)), 
  size=factor("5000"))
simsdf <- rbind(simsdf,sampledf)

sample2mean <- mean(simsdf$averages[simsdf$size == "5000"])

simsdf$mean[simsdf$size=="1000"] <- samplemean
simsdf$mean[simsdf$size=="5000"] <- sample2mean

library(ggplot2)
# plotting all the distributions with their means
ggplot(simsdf, aes(averages)) + 
  geom_histogram(aes(fill=size, color=size)) +
  geom_vline(aes(xintercept= mean)) +
  ggtitle("Expected Values of Various Sample Sizes")  +
  xlab("Averages") +
  facet_grid(.~size)
```

No matter the sample size, the sample mean approximates the theoretical mean. 

## Sample Variance versus Theoretical Variance 

Next, let's look at our sample variance. Since our mean and our standard deviation are the same, we know our theoretical variance is `r 1/.2^2/40` based on the following formula:

$$Var[X]=\frac{1}{\lambda^2}$$

Using the same formula, we can determine our sample variance.

```{r echo=FALSE, message=FALSE}

library(dplyr)
# subsetting the original simulation from my data frame
originalsim <- simsdf %>% filter(size=="1000")

samplevariance <- var(originalsim$averages)
samplevariance

```

The sample variance shares a similar spread with the theoretical variance.

## Distribution

There are three ways to tell if our distribution is normal. 

* mean = median = mode
* The distribution is symmetric
* 50% of values are less than the mean and 50% are greater
 
Two of these are easy to test.

### mean = median = mode.

In R, this is a simple logical comparison. Since we're trying to prove if it is "approximately" normal, let's use the signif function on our values to the first decimal point. (See Appendix D for code on generating the mode.)

```{r, echo=FALSE}

# table of original simulation to find max
table_data <- table(signif(originalsim$averages,1))

# mode of table_data
samplemode <- subset(table_data, table_data==max(table_data))
samplemode <- as.numeric(names(samplemode))

```


```{r}

signif(samplemean,1) == signif(median(originalsim$averages),1) 

signif(samplemean,1) == samplemode

```

By the transitive property of equality, if mean = median and mean = mode, then median= mode. In this case, that is true. 

### 50% of values are less than/above the mean

To find out if our values are split 50/50 at the mean, let's use the qexp function to find the 50th quartile.

```{r}

qexp(0.50,1/samplemean)

```

Uh-oh. If we use the pexp function on our mean, we find our distribution is actually split with `r pexp(samplemean, 1/samplemean, lower.tail=FALSE)*100`% of values occurring above the mean and `r pexp(samplemean, 1/samplemean)*100`% of values occuring below. That's hardly symmetric.

We're looking for "approximately" normal, not normal, however. While the distribution does not fit the strict definition of normal, one could argue that because it meets 1/3 of the criteria, it could be considered approximate.

\pagebreak

# Appendix

## Appendix A

### Simulation Script

```{r eval=FALSE}

nosims <- 1000
nexps <- 40
rate <- 0.2

# using set.seed so the results can be replicated
set.seed(13)

# 1000 simulations on 40 random exponentials
sims <- replicate(nosims,rexp(nexps,rate))

# storing the clt of those simulations in a data frame
# with the size of the simulations
simsdf <- data.frame(averages = apply(sims,2, mean),
                     size = factor("1000"))

```

## Appendix B

```{r echo=FALSE, message=FALSE, fig.height=4}

ggplot(originalsim, aes(averages)) + 
  geom_histogram() +
  geom_vline(xintercept= samplemean) +
  ggtitle("Sample Mean of Simulation") +
  xlab("Averages")

```

## Appendix C

### Running Additional Simulations and Adding Results to the Sims Data Frame

```{r eval=FALSE}

# simulate 5000 random exponentials
set.seed(13)
sims2 <- replicate(nosims*5,rexp(nexps,rate))

# creating a new data frame and then binding those rows to the original
sampledf <- data.frame(averages=c(
  apply(sims2,2, mean)), 
  size=factor("5000"))
simsdf <- rbind(simsdf,sampledf)

sample2mean <- mean(simsdf$averages[simsdf$size == "5000"])

simsdf$mean[simsdf$size=="1000"] <- samplemean
simsdf$mean[simsdf$size=="5000"] <- sample2mean

```

## Appendix D

### Calculating the Mode

```{r, eval=FALSE}

# subsetting the original simulation from my data frame
originalsim <- simsdf %>% filter(size=="1000")

# table of original simulation to find max
table_data <- table(signif(originalsim$averages,1))

# mode of table_data
samplemode <- subset(table_data, table_data==max(table_data))
samplemode <- as.numeric(names(samplemode))

```
